Lightweight Real-Time Feature Monitoring

### 2020-07-15 chapters/conclusions.tex
Experiment results for last test

\subsubsection*{Hypothesis Discussion}
In our synthetic data experiments, we were able to detect all of the introduced anomalies. Additionally, we were able to do it achieving very high streaming throughputs. In the real data experiments, once again, we managed to achieve high throughput but the produced alerts were not very accurate. We defer further investigation to future work because of the time constraints of this thesis but leave open the possibility that the approximate aggregation value is more inaccurate than what we thought it would be since the EMA tail holds all past events.

@@

    \item experiments on real data that showed unexpected results but where we provide insights and possible hypothesis to test for future work
\end{enumerate}

@@

First and foremost, we believe the next steps are to reason about the obtained results in the real data experiments and find the root cause for the existence of many and large alert periods. As a possible hypothesis to test, we suggest testing with EMAs with higher smoothing factors, so that past events are forgotten faster; additionally, we suggest decreasing $\gamma$, which further increases the number of samples to make in the batch phase.

@@

Lastly, we propose another type of validation method. We mentioned in the beginning that our objective is to detect data pattern shifts before the performance of the systems relying on the data declines. We propose a test where we train a Machine Learning model on a reference period and measure its prediction accuracy on a target period. Then, we propose running our method on reference and target periods to find out timestamps with a high number of divergent features. We propose that the model is retrained at these timestamps (tantamount to system reconfiguration) and then used to predict for the remaining target period. Finally, we propose that the accuracy should be compared to the previously observed model performance, expecting it to be better after retraining in one of the high divergent timestamps.

>>>>

\subsubsection*{Hypothesis Discussion}
In our synthetic data experiments, we were able to detect all of the introduced anomalies. Additionally, we were able to do it achieving very high streaming throughputs. In the real data experiments, once again, we managed to achieve high throughput but the produced alerts were not very accurate. We defer further investigation to future work because of the time constraints of this thesis but leave open the possibilities that our divergence measure distribution was poorly built and hence more samples need to be made in the batch phase or that the approximate EMA histogram aggregation value is more inaccurate than what we thought it would be since the EMA tail holds all past events and higher EMA smoothing factors should be used.

@@

    \item experiments on real data that showed unexpected results but where we provide insights and possible hypotheses to test for future work
\end{enumerate}

@@

First and foremost, we believe the next steps are to reason about the obtained results in the real data experiments and find the root cause for the existence of many and large alert periods. As a possible hypothesis to test, we suggest testing with EMAs with higher smoothing factors, so that past events are forgotten faster; additionally, we suggest decreasing $\gamma$, which further increases the number of samples to make in the batch phase, improving the accuracy of the divergence measure distribution built.

@@

Lastly, we propose another type of validation method. We mentioned in the beginning that our objective is to detect data pattern shifts before the performance of the systems that rely on the data declines. We propose a test where we train a Machine Learning model on a reference period and measure its prediction accuracy on a target period. Then, we propose running our method on reference and target periods to find out timestamps with a high number of divergent features. We propose that the model is retrained at these timestamps (tantamount to system reconfiguration) and then used to predict for the remaining target period. Finally, we propose that the accuracy should be compared to the previously observed model performance, expecting it to be higher after retraining in one of the high divergent timestamps.

### 2020-07-15 chapters/validation.tex
Final changes and submission
Experiment results for last test

\begin{table}[!htb]
    \begin{center}
        \begin{tabular}{|c|c|c|ll}
        \cline{1-3}
        \textbf{Feature} & \textbf{Reference Distribution} & \textbf{Parameters} &  &  \\ \cline{1-3}
        $x_1$               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
        $x_2$               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
        $x_3$               & Uniform                         & $a=200$, $b=250$       &  &  \\ \cline{1-3}
        $x_4$               & Uniform                         & $a=50$, $b=125$        &  &  \\ \cline{1-3}
        \end{tabular}
    \end{center}
    \caption{Multi feature reference dataset and feature distributions}
    \label{tbl:multi-feat-ref-dataset-distros}
\end{table}

@@

\section{Conclusions}
In this section, we summarize our experimental tests, results and uncovered issues.

We began by performing a single-feature analysis of synthetic data. The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported. Still working with synthetic data, we performed our first multi-feature analysis. In this scenario, we applied the Holm-Bonferroni multiple test correction to the p-values of each hypothesis. Once again, each feature was correctly reported as divergent at the timestamps where we changed the underlying generating distribution.

Finally, we moved on to real data. Due to time constraints and the scope of this thesis, we experimented only with data from one merchant. In our first experiment with real data, we ran our method with what we thought was an ideal set of parameters and produced a lot of alarms. A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as this dataset contained a lot of drift. In our last experiment, we attempted to reduce concept drift to a maximum by using the same period as reference and target. To our surprise, we still produced a large number of long-lasting alarms. We defer the investigation of the root cause of the problem to future work and leave a few untested hypotheses as solutions for this issue, like the possibility that the EMA-like histograms in batch work with subsets of data and the streaming ones work with infinite amounts of data. In other words, for the streaming EMAs, the previous events never have their contribution set to zero, even though the weights tend to zero, whereas the batch EMAs worked with a finite dataset. Additionally, our sampling may have been insufficient. To test this, an experiment with a lower $\gamma$ and hence a higher number of samples should be carried out.

>>>>

\begin{table}[!htb]
    \begin{center}
        \begin{tabular}{|c|c|c|ll}
        \cline{1-3}
        \textbf{Feature} & \textbf{Reference Distribution} & \textbf{Parameters} &  &  \\ \cline{1-3}
        $x_1$               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
        $x_2$               & Gaussian/Normal                 & $\mu=100$, $\sigma=5$  &  &  \\ \cline{1-3}
        $x_3$               & Uniform                         & $a=200$, $b=250$       &  &  \\ \cline{1-3}
        $x_4$               & Uniform                         & $a=50$, $b=125$        &  &  \\ \cline{1-3}
        \end{tabular}
    \end{center}
    \caption{Multi-feature reference dataset and feature distributions}
    \label{tbl:multi-feat-ref-dataset-distros}
\end{table}

@@

\subsection{Building an Accurate Divergence Measure Distribution}
Our method comprises two-phases: the batch and the streaming analysis. We can not conclude in which phase an error that could possibly explain the last experiment's results exists. To isolate both phases, we conduct this last experiment where instead of using the distribution of divergence measures built in the batch phase we use a distribution we know to be the best fit for the streaming period.

Our best fit divergence distribution was built over the streaming period. To do this, we processed the streaming events once and saved all the JSD measures. We binned these JSD values with 50,000 bins all with equal heights (or counts). Then we performed once again the streaming analysis over the same period but this time using the previously built distribution for which we know it is the best possible fit, since it was computed over the exact same period, built event by event. We set the FWER to be 1\% so we expect to see some alarms, \textit{i.e.}, for JSD measures above the 99th percentile.

Recall that in the last experiment, for feature $x_1$, using the same target and reference periods, we had two large alarms that lasted months (Figure \ref{fig:merchant2-x1-correctedpvalues}). Using the same periods but with the newly computed distribution of divergence measures we obtain the corrected p-values plot shown in Figure \ref{fig:merchant3-x1-correctedpvalues}.
\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/merchant3-x1-correctedpvalues.pdf}
      \caption{Merchant feature $x_1$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
      \label{fig:merchant3-x1-correctedpvalues}
    \end{center}
\end{figure}
This plot actually makes sense and represents expected behavior. Notice that we have only a few short-lasting alarms, which correspond to the JSD measures above the 99th percentile. These can be thought of as false positives. In reality, the definition of the multi-test we perform is exactly this: alert events above the 99th percentile (FWER=1\%).

Figures \ref{fig:merchant3-x2-correctedpvalues}, \ref{fig:merchant3-x3-correctedpvalues}, \ref{fig:merchant3-x4-correctedpvalues} and \ref{fig:merchant3-x5-correctedpvalues} show the corrected p-value plots for this experiment for features $x_2$, $x_3$, $x_4$ and $x_5$, respectively.
\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/merchant3-x2-correctedpvalues.pdf}
      \caption{Merchant feature $x_2$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
      \label{fig:merchant3-x2-correctedpvalues}
    \end{center}
\end{figure}
\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/merchant3-x3-correctedpvalues.pdf}
      \caption{Merchant feature $x_3$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
      \label{fig:merchant3-x3-correctedpvalues}
    \end{center}
\end{figure}
\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/merchant3-x4-correctedpvalues.pdf}
      \caption{Merchant feature $x_4$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
      \label{fig:merchant3-x4-correctedpvalues}
    \end{center}
\end{figure}
\begin{figure}[!htb]
    \begin{center}
      \includegraphics[scale=0.5]{figures/merchant3-x5-correctedpvalues.pdf}
      \caption{Merchant feature $x_5$ corrected p-values with equal reference and target periods and using the best-fit divergence distribution}
      \label{fig:merchant3-x5-correctedpvalues}
    \end{center}
\end{figure}
The analysis performed for feature $x_1$ applies for all of these features. Notice that for all plots we have very few alarms of very short durations. This coherency in results across multiple features seems to indicate that our streaming analysis is accurate and detects the outliers accurately if the divergence measure distribution used is well built.

\newpage
\section{Conclusions}
In this section, we summarize our experimental tests, results and uncovered issues.

We began by performing single-feature analysis of synthetic data. The results obtained with this set of experiments were what we expected, in the sense that the artificially introduced anomalies in the datasets were all correctly reported. Still working with synthetic data, we performed our first multi-feature analysis. In this scenario, we applied the Holm-Bonferroni multiple test correction to the p-values of each hypothesis. Once again, each feature was correctly reported as divergent at the timestamps where we changed the underlying generating distribution.

We then moved on to real data. Due to time constraints and the scope of this thesis, we experimented only with data from one merchant. In our first experiment with real data, we ran our method with what we thought was an ideal set of parameters and produced a lot of alarms. A close analysis of the reference and target time-series of each feature revealed that some alarms indeed made sense, as this dataset contained a lot of drift.

In our second experiment with real data, we attempted to reduce concept drift to a maximum by using the same period as reference and target. To our surprise, we still produced a large number of long-lasting alarms. We defer the investigation of the root cause of the problem to future work and leave a few untested hypotheses as solutions for this issue, like the possibility that the EMA-like histograms in batch work with subsets of data and the streaming ones work with infinite amounts of data. In other words, for the streaming EMAs, the previous events never have their contribution set to zero, even though the weights tend to zero, whereas the batch EMAs worked with a finite dataset. Additionally, our sampling may have been insufficient and hence the divergence measure distribution was inaccurate. To test the first hypothesis, experiments with higher smoothing factors for streaming EMAs should be carried out. To test the second one, a lower $\gamma$ parameter should be used in the batch phase, hence increasing the number of samples to be made, potentially building a more accurate distribution of JSD values.

In our final experiment, we aim to build the most accurate as possible distribution of divergence measures by building it based on the measured divergences observed in streaming. In our second streaming run, we used this distribution and observed that the alarms produced were very few and short-lasting ones. The few alarms raised correspond to divergence measures above the 99th percentile. With this experiment we conclude our streaming analysis is correctly built but depends a lot on how accurate the divergence measure distribution is. As discussed, future tests should be carried out where more samples are made, to ensure a more accurate divergence measure distribution is computed.
